Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2022-11-24 reports/supplement/supplement_0.9.1.Rmd
changes supplement

When SB contained multiple strains varying in growth rate (and tolerance) there were two stable states at intermediate levels of oxygen diffusivity, only one stable state at low oxygen diffusivity with only sulfur bacteria, and with only one stable state at high oxygen diffusivity with high cyanobacterial density (Figure \@ref(fig:max-tol2)), as was previously reported. When, however, only one SB strain with very high tolerance and low maximum growth rate was present, the stable states change: at low oxygen diffusivity  all three functional groups coexisted (Figure \@ref(fig:max-tol3)) – this never occurred when there were multiple SB strains varying in tolerance.

@@

(ref:max-tol2) Stable states found when there is variation among nine strains of SB and the amount of variation is high. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:max-tol3) Stable states found when SB contained only the most tolerant strain, with trait values corresponding with the diversity level of the previous figure. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

In the SB-CB case, the presence of only the most tolerant strain led to coexistence of all groups at low levels of oxygen diffusivity when trait values were moderately high (Figure \@ref(fig:max-tol55)) or to absence of alternate stable states and sole dominance of cyanobacteria when trait values were high (Figure \@ref(fig:max-tol5)). As discussed in the previous section, the low maximum growth rate of the most tolerant SB strain led to less production of sulfide such that SB could not exclude CB.

(ref:max-tol55)  Stable states found when CB and SB each contain only the single most tolerant strain with moderate tolerance. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:max-tol5)  Stable states found when CB and SB each contain only the single most tolerant strain with trait values corresponding with the diversity level of the previous figure. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:PB-stabstat4) Coexistence of all three groups at low oxygen diffusivities when only PB contains strain variation, and the amount of strain variation is moderate to high. This example is when the standardised amount of trait variation is 0.89 (0 being no variation, and 1 being the maximum amount in any simulation.)

@@

(ref:PB-stabstat5) Dynamics when there is high tolerance in only PB, with this being the same tolerance level as the most tolerant strain in Figure \@ref(fig:PB-stabstat4) immediately above.

@@

(ref:max-tol6) Stable states found with nine strains of CB-SB-PB and high diversity. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

(ref:max-tol7) Stable states found with the single most tolerant CB strain, SB strain, and PB strain, with trait values corresponding with the diversity level in the previous figure. All other aspects are the same as in Figure \@ref(fig:repl-meth2).

@@

When two functional groups varied in traits, the effects of trait variation were additive when one of the groups had no diversity effect on a tipping point but non-additive when both groups had diversity effects on a tipping point (Fig. \@ref(fig:fig-add-linear) a-f). High trait variation in cyanobacteria and sulfate-reducing bacteria caused the oxic-anoxic tipping point to occur at lower oxygen diffusivity than expected assuming additivity of the individual effects of diversity in these two groups (Fig. \@ref(fig:fig-add-linear)d). When the two groups of sulfur bacteria varied in traits, the anoxic-oxic tipping point occurred at higher oxygen diffusivity than expected by additivity (Fig. \@ref(fig:fig-add-linear)f).

When all three groups varied in traits, only the diversity effect in phototrophic sulfur bacteria on the oxic-anoxic tipping point was additive; all other effects were non-additive, especially at higher trait variation (Fig. \@ref(fig:fig-add-linear)g-i).

>>>>

When SB contained multiple strains varying in growth rate (and tolerance) there were two stable states at intermediate levels of oxygen diffusivity, only one stable state at low oxygen diffusivity with only sulfur bacteria, and only one stable state at high oxygen diffusivity with high cyanobacterial density (Figure \@ref(fig:max-tol2)), as was previously reported. When, however, only one SB strain with very high tolerance and low maximum growth rate was present, the stable states change: at low oxygen diffusivity  all three functional groups coexisted (Figure \@ref(fig:max-tol3)) – this never occurred when there were multiple SB strains varying in tolerance.

@@

(ref:max-tol2) Stable states found when there is variation among nine strains of SB and the amount of variation is high. All other aspects are the same as in Figure \@ref(fig:ss-temp-meth3).

@@

(ref:max-tol3) Stable states found when SB contained only the most tolerant strain, with trait values corresponding with the diversity level of the previous figure. All other aspects are the same as in Figure \@ref(fig:ss-temp-meth3).

@@

In the CB-SB case, the presence of only the most tolerant strain led to coexistence of all groups at low levels of oxygen diffusivity when trait values were moderately high (Figure \@ref(fig:max-tol55)) or to absence of alternate stable states and sole dominance of cyanobacteria when trait values were high (Figure \@ref(fig:max-tol5)). As discussed in the previous section, the low maximum growth rate of the most tolerant SB strain led to less production of sulfide such that SB could not exclude CB.

(ref:max-tol55)  Stable states found when CB and SB each contain only the single most tolerant strain with moderate tolerance. All other aspects are the same as in Figure \@ref(fig:ss-temp-meth3).

@@

(ref:max-tol5)  Stable states found when CB and SB each contain only the single most tolerant strain with high tolerance. All other aspects are the same as in Figure \@ref(fig:ss-temp-meth3).

@@

(ref:PB-stabstat4) Coexistence of all three groups at low oxygen diffusivities when only PB contains strain variation, and the amount of strain variation is medium to high. This example is when the standardised amount of trait variation is 0.89 (0 being no variation, and 1 being the maximum amount in any simulation.)

@@

(ref:PB-stabstat5) Dynamics when PB contain only the most tolerant strain, with this being the same tolerance level as the most tolerant strain in Figure \@ref(fig:PB-stabstat4) immediately above.

@@

(ref:max-tol6) Stable states found with nine strains of CB-SB-PB and high diversity. All other aspects are the same as in Figure \@ref(fig:ss-temp-meth3).

@@

(ref:max-tol7) Stable states found with the single most tolerant CB strain, SB strain, and PB strain, with trait values corresponding with the diversity level in the previous figure. All other aspects are the same as in Figure \@ref(fig:ss-temp-meth3).

@@
